---
title: "Multistate Modeling with a time-to event mediator Mediator"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Multistate Modeling with a time-to event mediator Mediator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This example demonstrates how to use `cmest` when there is a time-to-event mediator. For this purpose, we simulate several situation: 1) one baseline confounder. 2) multiple baseline confounders.  
Data contains a binary exposure $A$, a time-to-event mediator $M$, a time-to-event outcome $Y$ and multiple types of confounders $C^{'}$. 

We can define the Total Effect (TE) with respect to the survival function.
$$TE=P(S>s|A=1,C) - P(S>s|A=0,C)$$
The probability of surviving after time $s$ between $A=1$ and $A=0$ patients had the distribution of time to mediator $g$ been the same across the two exposure groups is referred as $"Residual\ Disparity" (RD)$
$$RD=P(S_g>s|A=1,C) - P(S_g>s|A=0,C)$$
The probability of surviving after time $s$ for subjects in exposure group $A=a$ for a change in the time to mediator level is referred as $"Shifting\ Distribution\ Effect" (SD)$
$$SDE=P(S_g>s|A=1,C) - P(S_{g^*}>s|A=1,C)$$

There are three transitions: $A$ $\rightarrow$ $M$, $A$ $\rightarrow$ $Y$ and $M$ $\rightarrow$ $Y$. So we define $\alpha_{01}(t|A,C)$, $\alpha_{02}(t|A,C)$ and $\alpha_{12}(t|A,C)$ as the instantaneous hazard of three transitions respectively. $\Lambda_{01}(s|A,C)$, $\Lambda_{02}(s|A,C)$ and $\Lambda_{12}(s|A,C)$ are the corresponding cumulative transition intensity functions.

We specify a semi-parametric proportional intensity model for the hazard of transitions:
$$\alpha_{01}(t|A,C)=\alpha_{01}^{0}(t)e^{\beta_1A+\beta_2^{'}C}$$

$$\alpha_{02}(t|A,C)=\alpha_{02}^{0}(t)e^{\gamma_1A+\gamma_2^{'}C}$$

$$\alpha_{12}(t|t^{'},A,C)=\alpha_{12}^{0}(t)e^{\delta_1A+\delta_2t^{'}+\delta_3A*t^{'}+\delta_4^{'}C}$$



  
  
```{r}
library(CMAverse)
```

The DAG for this scientific setting is:

```{r}
cmdag(outcome = "Y", exposure = "A", mediator = "M",
      basec = c("C1", "C2"), postc = NULL, node = TRUE, text_col = "white")
```

In this setting, we can use the multistate approach for stochastic interventions on a time-to event mediator in the presence of semi-competing risks. The results are shown below:

## The multistate approach for stochastic interventions on a time-to event mediator in the presence of semi-competing risks

### Binary exposure with multiple baseline confounders                                    

This example demonstrates how to use `cmest` when there are one binary exposure, categorical baseline confounder C1 and continuous basline confounder C2. For this purpose, we simulate some data containing a categorical baseline confounder $C_1$, a continuous basline confounder $C_2$,a binary exposure $A$, a time-to-event $M$ and a time-to-event outcome $Y$. The semi-parametric proportional intensity model for the hazard of 3 transitions are:

$$\alpha_{01}(t|A,C)=\alpha_{01}^{0}(t)e^{\beta_1A+\beta_2^{'}C}
= \alpha_{01}^{0}(t)e^{-0.3*A+0.4*C_1+0.5*C_2}, where\  \alpha_{01}^{0}(t) = \lambda_{01}\gamma_{01}t^{\gamma_{01}-1} = 0.1*0.8*t^{0.8-1}$$

$$\alpha_{02}(t|A,C)=\alpha_{02}^{0}(t)e^{\gamma_1A+\gamma_2^{'}C}
= \alpha_{02}^{0}(t)e^{0.2*A+0.3*C_1+0.4*C_2}, where\  \alpha_{02}^{0}(t) = \lambda_{02}\gamma_{02}t^{\gamma_{02}-1} = 0.07*0.12*t^{0.12-1}$$

$$\alpha_{12}(t|t^{'},A,C)=\alpha_{12}^{0}(t)e^{\delta_1A+\delta_2t^{'}+\delta_3A*t^{'}+\delta_4^{'}C}
= \alpha_{12}^{0}(t)e^{0.*4A+0*t^{'}+(-0.2)*A*t^{'}+0.5*C_1+0.6*C_2}, where\  \alpha_{12}^{0}(t) = \lambda_{12}\gamma_{12}t^{\gamma_{12}-1} = 0.1*0.5*t^{0.5-1}$$



```{r}
#generate dataset
set.seed(1)
#build a function to generate time-to-event data
gen_srv <- function(n, lambda, gamma, beta, X){
  X = as.matrix(X)
  beta = as.matrix(beta, ncol=1)
  T = (-log(runif(n)) / (lambda * exp(X %*% beta)))^(1/gamma) #weibull distribution
  return(T)
}
n <- 5000
A = sample(c(0,1),replace=TRUE, size=n, c(0.7,0.3)) #binary exposure
c1 = sample(c(0,1),replace=TRUE, size=n,c(0.6, 0.4)) #binary confounder
c2 = rnorm(n, mean = 1, sd = 1) #continuous confounder
id=c(1:n)
full = data.frame(id,A,c1,c2)
M = gen_srv(n=n, lambda = 0.1,gamma = 0.8, beta = c(-0.3,0.4,0.5), X=data.frame(A,c1,c2)) #time to event mediator
Y = gen_srv(n=n, lambda = 0.07, gamma = 0.12, beta = c(0.2,0.3,0.4), X=data.frame(A,c1,c2)) #time to event outcome
data = data.frame(id = c(1:n), M = M, Y = Y)
# indicator for event
data$M_ind = ifelse(data$M <= data$Y, 1, 0)
data$Y_ind = 1
data <- merge(data,full , by = "id")
#modify Y distribution
trans_matrix = transMat(x = list(c(2, 3), c(3), c()), names = c("A", "M", "Y"))
covs = c("A","M", "c1","c2")
pre_data = msprep(time = c(NA, "M", "Y"), status = c(NA, "M_ind", "Y_ind"),
                         data = data, trans = trans_matrix, keep = covs)
pre_data = expand.covs(pre_data, covs, append = TRUE, longnames = FALSE)
pre_data$A_M.3 = pre_data$A.3*pre_data$M.3
# resample for T < S
data_23= pre_data[which(pre_data$trans == 3),]
data_23_tem = data.frame(id = rep(NA,dim(data_23)[1]),
                          new_y = rep(NA,dim(data_23)[1]))

for(i in 1:dim(data_23)[1]){
  data_23_tem$id[i] = data_23$id[i]
  repeat {
    # do something
    time_test = gen_srv(n = 1, 
                        lambda = 0.1,  
                        gamma = 0.5,
                        beta = c(as.numeric(0.4),
                                 0,
                                 as.numeric(0.5),
                                 as.numeric(0.6),
                                 as.numeric(-0.2)), 
                        X = data_23[i, c("A.3", "M.3", "c1.3","c2.3", "A_M.3")])
    # exit if the condition is met
    if (time_test > data_23[i,"M.3"]) break
  }
  data_23_tem$new_y[i] = time_test
}
data_temp = merge(data, data_23_tem, by = "id", all = T)
#modify Y and M
data_temp$Y[which(data_temp$M_ind == 1)] = data_temp$new_y[which(data_temp$M_ind == 1)]
data_temp$M[which(data_temp$M_ind == 0)] = data_temp$Y[which(data_temp$M_ind == 0)]
data_final = data_temp
data_final$Y_day = data_final$Y*30
data_final$M_day = data_final$M*30
data_final$Y_ind[which(data_final$Y > 24)] = 0 #censored data
data_final$Y[which(data_final$Y> 24)] = 24
data_final$Y_day[which(data_final$sY_day > 24*30)] = 24*30
data_final$M_ind[which(data_final$M > 24)] = 0
data_final$M[which(data_final$M > 24)] = 24
data_final$M_day[which(data_final$M_day > 24*30)] = 24*30
data_final$A = as.factor(data_final$A) #generate a factor exposure

data = data_final %>% select(id,A,M,Y,M_ind,Y_ind,c1,c2)

```




```{r message=F,warning=F,results='hide'}
surv_biE <- cmest(data = data, model = 'multistate',total_duration = 24, time_grid = 1,survival_time_fortable = 22, exposure = 'A',mediator = 'M', outcome = 'Y',
      event = "Y_ind",mediator_event = "M_ind", basec = c('c1','c2'),
      basecval = c('c1' = '0','c2' = '0'),astar = '0',a='1',nboot=100)
```

```{r message=F,warning=F}
summary(surv_biE)
```

```{r message=F,warning=F}
ggcmest(surv_biE)
```



### Categorical exposure with multiple baseline confounders                                    

This example demonstrates how to use `cmest` when there are one categorical exposure, categorical baseline confounder C1 and continuous basline confounder C2. For this purpose, we simulate some data containing a categorical baseline confounder $C_1$, a continuous basline confounder $C_2$,a categorical exposure $A$, a time-to-event $M$ and a time-to-event outcome $Y$. The semi-parametric proportional intensity model for the hazard of 3 transitions are:

$$\alpha_{01}(t|A,C)=\alpha_{01}^{0}(t)e^{\beta_1A+\beta_2^{'}C}
= \alpha_{01}^{0}(t)e^{-0.3*A_1+(-0.2*A_2)+0.4*C_1+0.5*C_2}, where\  \alpha_{01}^{0}(t) = \lambda_{01}\gamma_{01}t^{\gamma_{01}-1} = 0.1*0.8*t^{0.8-1}$$

$$\alpha_{02}(t|A,C)=\alpha_{02}^{0}(t)e^{\gamma_1A+\gamma_2^{'}C}
= \alpha_{02}^{0}(t)e^{0.2*A_1+0.3*A_2+0.3*C_1+0.4*C_2}, where\  \alpha_{02}^{0}(t) = \lambda_{02}\gamma_{02}t^{\gamma_{02}-1} = 0.07*0.12*t^{0.12-1}$$

$$\alpha_{12}(t|t^{'},A,C)=\alpha_{12}^{0}(t)e^{\delta_1A+\delta_2t^{'}+\delta_3A*t^{'}+\delta_4^{'}C}
= \alpha_{12}^{0}(t)e^{0.4*A_1+0.3*A_2+0*t^{'}+(-0.2)*A_1*t^{'}+(-0.3)*A_2*t^{'}+0.5*C_1+0.6*C_2}, where\  \alpha_{12}^{0}(t) = \lambda_{12}\gamma_{12}t^{\gamma_{12}-1} = 0.1*0.5*t^{0.5-1}$$


```{r}
#generate dataset
set.seed(1)
#build a function to generate time-to-event data
gen_srv <- function(n, lambda, gamma, beta, X){
  X = as.matrix(X)
  beta = as.matrix(beta, ncol=1)
  T = (-log(runif(n)) / (lambda * exp(X %*% beta)))^(1/gamma) #weibull distribution
  return(T)
}
n <- 10000
A = sample( c(0,1,2), size = n, replace=TRUE, prob=c(0.3,0.4,0.3))#categorical exposure
A=as.factor(A)
c1 = sample(c(0,1),replace=TRUE, size=n,c(0.6, 0.4)) #binary confounder
c2 = rnorm(n, mean = 1, sd = 1) #continuous confounder
id=c(1:n)
full = data.frame(id,A,c1,c2)
A1 = ifelse(full$A == 1, 1, 0)
A2 = ifelse(full$A == 2, 1, 0)
M = gen_srv(n=n, lambda = 0.1,gamma = 0.8, beta = c(-0.3,-0.2,0.4,0.5), X=data.frame(A1,A2,c1,c2)) #time to event mediator
Y = gen_srv(n=n, lambda = 0.07, gamma = 0.12, beta = c(0.2,0.3,0.4,0.5), X=data.frame(A1,A2,c1,c2)) #time to event outcome
data = data.frame(id = c(1:n), M = M, Y = Y)
# indicator for event
data$M_ind = ifelse(data$M <= data$Y, 1, 0)
data$Y_ind = 1
data <- merge(data,full , by = "id")
#modify Y distribution
trans_matrix = transMat(x = list(c(2, 3), c(3), c()), names = c("A", "M", "Y"))
covs = c("A","M", "c1","c2")
pre_data = msprep(time = c(NA, "M", "Y"), status = c(NA, "M_ind", "Y_ind"),
                  data = data, trans = trans_matrix, keep = covs)
pre_data = expand.covs(pre_data, covs, append = TRUE, longnames = FALSE)
pre_data$A1_M.3 = pre_data$A1.3*pre_data$M.3
pre_data$A2_M.3 = pre_data$A2.3*pre_data$M.3
# resample for T < S
data_23= pre_data[which(pre_data$trans == 3),]
data_23_tem = data.frame(id = rep(NA,dim(data_23)[1]),
                         new_y = rep(NA,dim(data_23)[1]))

for(i in 1:dim(data_23)[1]){
  data_23_tem$id[i] = data_23$id[i]
  repeat {
    # do something
    time_test = gen_srv(n = 1, 
                        lambda = 0.1,  
                        gamma = 0.5,
                        beta = c(as.numeric(0.4),
                                 as.numeric(0.3),
                                 0,
                                 as.numeric(0.5),
                                 as.numeric(0.6),
                                 as.numeric(-0.2),
                                 as.numeric(-0.3)), 
                        X = data_23[i, c("A1.3","A2.3", "M.3", "c1.3","c2.3", "A1_M.3","A2_M.3")])
    # exit if the condition is met
    if (time_test > data_23[i,"M.3"]) break
  }
  data_23_tem$new_y[i] = time_test
}
data_temp = merge(data, data_23_tem, by = "id", all = T)
#modify Y and M
data_temp$Y[which(data_temp$M_ind == 1)] = data_temp$new_y[which(data_temp$M_ind == 1)]
data_temp$M[which(data_temp$M_ind == 0)] = data_temp$Y[which(data_temp$M_ind == 0)]
data_final = data_temp
data_final$Y_day = data_final$Y*30
data_final$M_day = data_final$M*30
data_final$Y_ind[which(data_final$Y > 24)] = 0 #censored data
data_final$Y[which(data_final$Y> 24)] = 24
data_final$Y_day[which(data_final$sY_day > 24*30)] = 24*30
data_final$M_ind[which(data_final$M > 24)] = 0
data_final$M[which(data_final$M > 24)] = 24
data_final$M_day[which(data_final$M_day > 24*30)] = 24*30
#data_final$A = as.factor(data_final$A) #generate a factor exposure

data_catE = data_final %>% select(id,A,M,Y,M_ind,Y_ind,c1,c2)


```




```{r message=F,warning=F,results='hide'}
surv_catE <- cmest(data = data_catE, model = 'multistate',total_duration = 24, time_grid = 1,survival_time_fortable = 22, exposure = 'A',mediator = 'M', outcome = 'Y',
      event = "Y_ind",mediator_event = "M_ind", basec = c('c1','c2'),
      basecval = c('c1' = '0','c2' = '0'),astar = '0',a='2',nboot=100)
```

```{r message=F,warning=F}
summary(surv_catE)
```

```{r message=F,warning=F}
ggcmest(surv_catE)
```
